% !TEX root = master.tex


\section*{Preamble}
\addcontentsline{toc}{section}{Preamble}

\begin{solution}
\emph{Disclaimer:} The reference solutions in the following range from exact results to sketch solutions; this is entirely on purpose. Some of the assignments are very open, which reflects to our solutions being just one of many. Of course we try to give a good solution and even solutions that are much more detailed than what we expect you to give. On the other hand, you would expect to give more detailed solutions that our sketches. Given the broad spectrum of solutions, it is not possible to directly infer any grade level from this document. \textcolor{solutioncolor}{All solutions have been written in the in-lined space with this colour.}
\end{solution}

This is the text is an excerpt of the exam set for the 4 hour written re-exam in Computer Systems (CompSys), B1+2-2017/18.
This document consists of \pageref{LastPage} pages excluding this preamble; make sure you have them all. Read the rest of this preamble carefully. 
Your submission will be graded as a whole, on the 7-point grading scale, with external censorship.

\begin{itemize}
  \item You can answer in either Danish or English.
  \item Remember to write your exam number on all pages.
  \item You do not have to hand-in this preamble.
\end{itemize}

\subsection*{Expected usage of time and space}
The set is divided into sub-parts that each are given a rough guiding estimate of the time needed. However, your exact usage of time can differ depending on prior knowledge and skill.

Furthermore, all questions includes formatted space (lines, figures, tables, etc.) for in-line answers. Please use these as much as possible. The available spaces are intended to be large enough to include a satisfactory answer of the question; thus, full answers of the question does not necessarily use all available space.
\\
If you find yourself in a position where you need more space or have to redo (partly) an answer to a question, continue on the backside of a paper or write on a separate sheet of paper. Ensure that the question number is included and that you in the in-lined answer space refers to it; e.g. write ``\emph{The [rest of this] answer is written on backside of/in appended page XX.}''

For the true/false and multiple-choice questions with one right answer give only one clearly marked answer. If more answers are given, it will be interpreted as incorrectly answered. Thus, if you change your answer, make sure that this shows clearly.


\subsection*{Exam Policy}
This is an \emph{individual}, open-book exam.
You may use the course book, notes and any documents printed or stored on your computer, but you may not search the Internet or communicate with others to answer the exam.

\subsection*{Errors and Ambiguities}

In the event of errors or ambiguities in the exam text, you are expected to
state your assumptions as to the intended meaning in your answer.
%
Some ambiguities may be intentional.


